Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2010-11-18 rjournal/RJtemplate.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@29 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{
Methodology extending nonparametric goodness-of-fit tests
to discrete null distributions has existed
for several decades. Despite this, modern statistical software
has generally failed to provide this methodology
to users. We offer a revision of R's \code{ks.test()} function
and a new \code{cvm.test()} function that serve to fill this need
in the R language for two of the most popular nonparametric
goodness-of-fit tests. This paper describes these contributions and
provides examples of their usage. Particular attention is given
to various numerical issues that arise in their implementation.
}

@@

%General idea of goodness-of-fit tests (hypothesis test about underlying distribution).
%Given an observed sequence of data, the typical question of statistical inference is to determine
%the underlying process from which it came. 
Goodness-of-fit tests are used to assess whether data are consistent
with a hypothesized null distribution.  The $\chi^2$ test is the best-known
parametric goodness-of-fit test, while the most popular nonparametric tests
are the classic test proposed by Kolmogorov and Smirnov followed closely by
several variants on a procedure proposed by
Cram\'{e}r and von Mises (tests sometimes called Cram\'{e}r-von Mises-Smirnov
tests, or simply Cram\'{e}r-von Mises tests) .

@@

In their most basic forms, these nonparametric goodness-of-fit
tests are intended for continuous null distributions, but they have
also been adapted for use with discrete null distributions.  Unfortunately,
most modern statistical software have failed to incorporate these discrete
versions, leaving researchers with the $\chi^2$ test or in the unfortunate
position of applying a nonparametric test designed for a continuous null
distribution and hoping for the best.
As we will see, this latter course of action can be particularly dangerous
with small sample sizes. This paper presents a revision of R's
\code{ks.test()} function and a new \code{cvm.test()} function to
fill this void for researchers and practitioners in the R environment.

@@

The most popular nonparametric goodness-of-fit test is the
Kolmogorov-Smirnov test. The idea is fairly simple:
Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution
and the empirical distribution function $F_{data}(x)$ of the observed data,  the test statistic is given by
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|.
\end{align}
The distribution of $D$ does not depend on the hypothesized
distribution, making this a computationally
attractive method. \cite{slakter65} offers a standard presentation
of the test and its
performance relative to other algorithms. 
The test statistic (1) is easily adapted for one-sided tests.
For these, the absolute value is discarded and the tests are based
on either the supremum of the remaining difference (the `greater' testing
alternative) or by replacing the supremum with a negative infimum
(the `lesser' hypothesis alternative).  Tabulated p-values have been
available for these tests since 1933. %(*** year ****).

@@

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions does not have such a clean solution. The formula of
the test statistic $D$ remains unchanged, but the distribution
of the test statistic is much more difficult to obtain. Unlike the 
continuous case, the null distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
\citep{stephens1974}.  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing exact p-values
in the case of discrete null distributions.

@@

The implementation of the discrete Kolmogorov-Smirnov test is considered
in two parts. First, the particular test statistic needs to be calculated
(corresponding to the desired one-sided or two-sided test).
Then, the p-value for that particular test statistic may be computed.

The test statistic is, in terms of the theory, the same as in the
continuous case; it would seem that no additional work would be required,
but unfortunately this is not the case.
Consider two non-decreasing functions $f$ and $g$, where the function $f$ is a step function with jumps on the set $\{x_1, \ldots x_N \}$ and $g$
is continuous (the classical Kolmogorov-Smirnov situation).
In order to determine the supremum of the difference between
these two functions, notice that
\begin{align*}
&\sup_x \left| f(x)- g(x) \right|  \nonumber  \\
       &= \max_i \left( \max\left( \left|g(x_i) - f(x_i) \right|, \lim_{x \rightarrow x_i} \left| g(x) - f(x_{i-1}) \right| \right)\right) \\
&=  \max_i \bigg(\max \bigg( \left|g(x_i) - f(x_i) \right|, \left| g(x_i) - f(x_{i-1}) \right| \bigg) \bigg).
\end{align*}
Computing the maximum over these $2N$
values (with $f$ equal to
$F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic for
a continuous null distribution. When the function $g$ is not
continuous, however, equality (3) does not hold in general because 
we cannot replace $\lim_{x\rightarrow x_e} g(x)$ with the value $g(x_i)$.

If it is known that $g$ is a step function, it follows that
for some small $\epsilon$,
\begin{align}
\sup_x &\left| f(x)- g(x) \right| =  \nonumber \\
        &                    \max_i \{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i - \epsilon) - f(x_{i-1}) \right| \}
\end{align}
where the discontinuities in $g$ are at least a distance $\epsilon$ apart. 
This, however, requires knowledge that $g$ is a step function as well as of
the nature of its support (or break-points).
%
%*** Hmm... do we care about time? ***  When the break-points of $g$ are
%unknown, or when $g$ is neither a step function nor a continuous distribution
%function, computing the supremum requires taking the numerical limit in (1***?). 
%This is computational involved and risks problems with numerical tolerances.
%
We implement the Kolmogorov-Smirnov test statistic for discrete null
distributions by forcing the user to completely specify the null distribution.

@@

Having obtained the test statistic, the p-value must then be calculated. 
For larger sample sizes, the null distribution can be approximated
with the standard null distribution. 
When an exact p-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used. 
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in Conover's original paper.

@@

While the Kolmogorov-Smirnov test may be the most popular of
the nonparametric goodness-of-fit tests, there is another family of 
tests which has been shown to be more powerful against a large class
of alternatives hypotheses. 
The original test was developed by
Harald Cram\'{e}r and Richard von Mises \citep{cramer1928, vonmises1928} and further adapted by \cite{anderson1952}, and  \cite{Watson1961}. 
The original test statistic, $W^2$, Anderson's $A^2$, and Watson's
$U^2$ are:
\begin{align}
W^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) \right]^2 dF_0(x) \\
A^2 &= n \cdot \int_{-\infty}^{\infty} \frac{\left[F_{data}(x)- F_{0}(x) \right]^2}{F_0(x) -F_0(x)^2} dF_0(x) \\
U^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) - W^2 \right]^2 dF_0(x)
\end{align}
As with the original Kolmogorov-Smirnov test statistic, these all have 
null distributions of their test statistics which are independent of the
specified null models.

The relative powers of these tests to different alternatives are
studied in depth in \cite{stephens1974}. The $W^2$
statistic was the original test statistic; it has the benefit of having
the simplest form.
The $A^2$ statistic was developed by
Anderson in the process of generalizing the test for the two-sample case.
Watson's $U^2$ statistic was developed for distributions
which are cyclic (with an ordering to the support but
no natural starting point), as it is invariant to cyclic reordering of
the support. 
For example, a distribution on the months of the year could be considered
cyclic.

It has been shown that these tests can be more powerful than Kolmogorov-Smirnov tests to certain deviations. As they all involve
integration over the whole range of data, rather than one supremum, it is not surprising that they are generally best when the
true alternative distribution deviates a little over the whole range of data rather than deviating a lot over a small range. For
a complete analysis of the relative powers of these tests see \cite{stephens1974}.

@@

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, although the discreteness allows for slightly simpler representations.
The null distribution of the test statistics are again hypothesis-dependent. The methods do not offer finite sample results, but rather 
show that the asymptotic distribution of the test statistic under the null hypothesis is equal to the distribution of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). 
We implement this asymptotic distribution here;
the original papers %(*** really, all of them? ***)
show that this approximation is conservative 
and asymptotically equivalent to the true null distribution.

%*** What does it mean for a distribution to be equal to a weighted sum of
%independent chi-squared variables?  I think this needs to be stated more
%precisely, above.  A formal definition of $Q$ is needed someplace, too,
%as I'll probably note below. ***

@@

Calculation of the three test statistics is done using the
matrix algebra given by \cite{choulakian1994}. The form of the
asymptotic null distribution is obtained using R's
eigenvalue decomposition functions.

The only notable difficulty in the process involves calculating the percentiles
of the weighted sum of chi-squares,
\begin{align}
Q = \sum_{i=1}^{p} \lambda_i \chi^2_{i,1df}
\end{align}
\cite{imhof1961} provides a method for obtaining the distribution of
a weighted sum of independent chi-squared variables.  It is
easily adapted for our case because
the chi-squared variables have only one degree of freedom.
The exact formula given for the distribution function of $Q$
is given by
\begin{align}
\mathbb{P}\{Q \geq x \} = \frac{1}{2} + 
\frac{1}{\pi} \int_{0}^{\infty} \frac{\sin\theta(u,x)}{u \rho(u) } du
\end{align}
for continuous functions $\theta(\cdot, x)$ and $\rho(\cdot)$ depending on the actual weights used.

There is no analytic solution to the integral, so the integration is
accomplished numerically. This seems fine in most situations we considered,
but numerical issues appear in the regime of large test statistics $x$
(or, equivalently, small p-values).
The function $\theta(\cdot, x)$ is linear in $x$; as the test statistic 
grows the corresponding periodicity of the integrand decreases and
the approximation becomes unstable. 
%This is further magnified in situations where the 
%$p$-values should be very small, where tiny fluctuations (undetectable
%elsewhere) become quite prominent. 
We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this approximation is useful for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation); it should not greatly affect the interpretation of the results.

%*** How is the cutoff established in the code? ***

Figure \ref{cvmissues} shows the non-monotonicity of the numerical
evaluation of equation 7 for the null distribution which is
uniform on the set $\{1,2,3\}$, as well as the behavior of the
conservative approximation in equation 9.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of p-value for given test statistics using numerical integration
(pink) versus a conservative chi-squared bound (dashed blue). The null distribution
is unform on the set $\{1,2,3\}$. }
\label{cvmissues}
\end{figure}

>>>>

\abstract{
Methodology extending nonparametric goodness-of-fit tests
to discrete null distributions has existed
for several decades. However, modern statistical software
has generally failed to provide this methodology
to users. We offer a revision of R's \code{ks.test()} function
and a new \code{cvm.test()} function that fill this need
in the R language for two of the most popular nonparametric
goodness-of-fit tests. This paper describes these contributions and
provides examples of their usage. Particular attention is given
to various numerical issues that arise in their implementation.
}

@@

%General idea of goodness-of-fit tests (hypothesis test about underlying distribution).
%Given an observed sequence of data, the typical question of statistical inference is to determine
%the underlying process from which it came. 
Goodness-of-fit tests are used to assess whether data are consistent
with a hypothesized null distribution.  The $\chi^2$ test is the best-known
parametric goodness-of-fit test, while the most popular nonparametric tests
are the classic test proposed by Kolmogorov and Smirnov followed closely by
several variants on Cram\'{e}r-von Mises tests.

%procedure proposed by
%Cram\'{e}r and von Mises (tests sometimes called Cram\'{e}r-von Mises-Smirnov
%tests, or simply Cram\'{e}r-von Mises tests).

@@

In their most basic forms, these nonparametric goodness-of-fit
tests are intended for continuous null distributions, but they have
also been adapted for discrete null distributions.  Unfortunately,
most modern statistical software packages and programming environments
have failed to incorporate these discrete versions.  As a result,
researchers would typically rely upon the $\chi^2$ test or
use of a nonparametric test designed for a continuous null distribution.
For smaller sample sizes, in particular, both of these choices can produce
misleading inferences.

This paper presents a revision of R's
\code{ks.test()} function and a new \code{cvm.test()} function to
fill this void for researchers and practitioners in the R environment. 
We first present overviews of the theory and general implementation of the
discrete Kolmogorov-Smirnov and Cram\'{e}r-von Mises tests.  We discuss
the particular implementation of tests in R and provide examples.  We
conclude with a short discussion, including the state of existing continuous
and two-sample Cram\'{e}r-von Mises testing in R.

@@

The most popular nonparametric goodness-of-fit test is the
Kolmogorov-Smirnov test.
Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution
and the empirical distribution function $F_{data}(x)$ of the observed data,  the test statistic is given by
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|.    \label{eqD}
\end{align}
The distribution of $D$ does not depend on the hypothesized
distribution, making this a computationally
attractive method. \cite{slakter65} offers a standard presentation
of the test and its
performance relative to other algorithms. 
The test statistic is easily adapted for one-sided tests.
For these, the absolute value in (\ref{eqD}) is discarded and the tests are based
on either the supremum of the remaining difference (the `greater' testing
alternative) or by replacing the supremum with a negative infimum
(the `lesser' hypothesis alternative).  Tabulated p-values have been
available for these tests since 1933. %(*** year ****).

@@

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions is not straightforward. The formula of
the test statistic $D$ remains unchanged, but its distribution
is much more difficult to obtain; unlike the 
continuous case, its distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
(see *** or *** or \citet{stephens1974}, for example).  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing exact p-values
in the case of discrete null distributions.

@@

The implementation of the discrete Kolmogorov-Smirnov test involves
two steps. First, the particular test statistic is calculated
(corresponding to the desired one-sided or two-sided test).
Then, the p-value for that particular test statistic may be computed.

The form of the test statistic is the same as in the
continuous case; it would seem that no additional work would be required
for the implemenetation, but this is not the case.
Consider two non-decreasing functions $f$ and $g$, where the function $f$ is a step function with jumps on the set $\{x_1, \ldots x_N \}$ and $g$
is continuous (the classical Kolmogorov-Smirnov situation).
In order to determine the supremum of the difference between
these two functions, notice that **** FORMAT THESE AT END ****
\begin{align}
&\sup_x \left| f(x)- g(x) \right|  \nonumber  \\
       &= \max_i \left( \max\left\{ \left|g(x_i) - f(x_i) \right|, \lim_{x \rightarrow x_i} \left| g(x) - f(x_{i-1}) \right| \right\} \right) \label{SUP1}\\
&=  \max_i \bigg(\max \bigg\{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i) - f(x_{i-1}) \right| \bigg\} \bigg). \label{SUP2}
\end{align}
Computing the maximum over these $2N$
values (with $f$ equal to
$F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic for
a continuous null distribution. When the function $g$ is not
continuous, however, equality (\ref{SUP2}) does not hold in general because 
we cannot replace $\lim_{x\rightarrow x_i} g(x)$ with the value $g(x_i)$.

If it is known that $g$ is a step function, it follows that
for some small $\epsilon$,
\begin{align}
\sup_x &\left| f(x)- g(x) \right| =  \nonumber \\
        &                    \max_i \{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i - \epsilon) - f(x_{i-1}) \right| \} \label{epsilon}
\end{align}
where the discontinuities in $g$ are more than some distance $\epsilon$ apart. 
This, however, requires knowledge that $g$ is a step function as well as of
the nature of its support (or break-points).
%
%*** Hmm... do we care about time? ***  When the break-points of $g$ are
%unknown, or when $g$ is neither a step function nor a continuous distribution
%function, computing the supremum requires taking the numerical limit in (1***?). 
%This is computational involved and risks problems with numerical tolerances.
%
We implement the Kolmogorov-Smirnov test statistic for discrete null
distributions by forcing the user to completely specify the null distribution.

@@

Having obtained the test statistic, the p-value must then be calculated. 
For larger sample sizes, the null distribution can be approximated
with the null distribution from the classical Kolmogorov-Smirnov test. 
When an exact p-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used. 
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in Conover's original paper.

@@

While the Kolmogorov-Smirnov test may be the most popular of
the nonparametric goodness-of-fit tests, there is another family of 
tests which has been shown to be more powerful against a large class
of alternatives hypotheses. 
The original test was developed by
Harald Cram\'{e}r and Richard von Mises \citep{cramer1928, vonmises1928} and further adapted by \cite{anderson1952}, and  \cite{Watson1961}. 
The original test statistic, $W^2$, Anderson's $A^2$, and Watson's
$U^2$ are:
\begin{align}
W^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) \right]^2 dF_0(x) \label{W2} \\
A^2 &= n \cdot \int_{-\infty}^{\infty} \frac{\left[F_{data}(x)- F_{0}(x) \right]^2}{F_0(x) -F_0(x)^2} dF_0(x) \label{A2} \\
U^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) - W^2 \right]^2 dF_0(x) \label{U2}
\end{align}
As with the original Kolmogorov-Smirnov test statistic, these all have 
null distributions of their test statistics which are independent of the
specified null models.

The $W^2$
statistic was the original test statistic; it has the benefit of having
the simplest form.
The $A^2$ statistic was developed by
Anderson in the process of generalizing the test for the two-sample case.
Watson's $U^2$ statistic was developed for distributions
which are cyclic (with an ordering to the support but
no natural starting point), as it is invariant to cyclic reordering of
the support. 
For example, a distribution on the months of the year could be considered
cyclic.

It has been shown that these tests can be more powerful than Kolmogorov-Smirnov tests to certain deviations from the hypothesized distribution. 
As they all involve
integration over the whole range of data, rather than use of a supremum.
Thus, they are best-suited for situations where the
true alternative distribution deviates a little over the whole support
rather than having large deviations over a small section of the support. For
a complete analysis of the relative powers of these tests, see 
\cite{stephens1974}.

@@

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, 
and the null distributions of the test statistics are again hypothesis-dependent. The methods do not offer finite-sample results, but rather 
show that the asymptotic distributions of the test statistics under the null hypothesis each involve consideration of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). 
We implement the asymptotic distributions here; \cite{stephens1974}
showed that these approximations are conservative 
and asymptotically equivalent to the true null distributions.

@@

Calculation of the three test statistics is done using the
matrix algebra given by \cite{choulakian1994}. 
The only notable difficulty in the implementation of the discrete form
of the tests involves calculating the percentiles
of the weighted sum of chi-squares,
\begin{align}
Q = \sum_{i=1}^{p} \lambda_i \chi^2_{i,1df},   \label{eqQ}
\end{align}
where $p$ is the number of elements in the support of the hypothesized
distribution.
\cite{imhof1961} provides a method for obtaining the distribution of $Q$,
easily adapted for our case because
the chi-squared variables have only one degree of freedom.
The exact formula given for the distribution function of $Q$
is given by
\begin{align}
\mathbb{P}\{Q \geq x \} = \frac{1}{2} + 
\frac{1}{\pi} \int_{0}^{\infty} \frac{\sin\theta(u,x)}{u \rho(u) } du
\label{distQ}
\end{align}
for continuous functions $\theta(\cdot, x)$ and $\rho(\cdot)$ depending on the  weights $\lambda_i$.

There is no analytic solution to the integral in (\ref{distQ}), 
so the integration is
accomplished numerically. This seems fine in most situations we considered,
but numerical issues appear in the regime of large test statistics $x$
(or, equivalently, small p-values).
The function $\theta(\cdot, x)$ is linear in $x$; as the test statistic 
grows the corresponding periodicity of the integrand decreases and
the approximation becomes unstable. 
%This is further magnified in situations where the 
%$p$-values should be very small, where tiny fluctuations (undetectable
%elsewhere) become quite prominent. 
We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this approximation is used for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).

Figure \ref{cvmissues} shows the non-monotonicity of the numerical
evaluation of equation (\ref{distQ}) for the null distribution which is
uniform on the set $\{1,2,3\}$, as well as the behavior of the
conservative upper bound given in equation (\ref{ineq2}).

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of calculated p-values for given test statistics
using numerical integration
(pink) compared to the conservative chi-squared bound (dashed blue). 
The null distribution is uniform on the set $\{1,2,3\}$ in this example.
The instability in the calculated p-values is a result of numerical
instabilities.}
\label{cvmissues}
\end{figure}
